Random Matrix Theory Tools for the Predictive Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-15 paper/paper/SPIE/article.tex
finalize
tidy up and make corrections to paper

\title{Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations}

\author[a]{Derek Berger}
\author[a,*]{Jacob Levman}
\author[b]{Gurpreet M. Matharoo}
\affil[a]{St. Francis Xavier University, Department of Computer Science, 4130 University Avenue, Antigonish, Canada, B2G 2W5}
\affil[b]{St. Francis Xavier University, ACENET, 4130 University Avenue, Antigonish, Canada, B2G 2W5}

@@

% \begin{abstract} This document shows the required format and appearance of a
% manuscript prepared for SPIE journals. It is prepared using LaTeX2e with the
% class file \texttt{spieman.cls}. Please note that the following journals
% require the use of structured abstracts in manuscript submissions:
% \textit{Neurophotonics}, the \textit{Journal of Biomedical Optics}, and the
% \textit{Journal of Medical Imaging}. Structured abstracts are encouraged for
% the \textit{Journal of Micro/Nanolithography, MEMS, and MOEMS}. Guidelines
% are available on the journal website. Whether structured or single-paragraph,
% the abstract should be a summary of the paper and not an introduction.
% Because the abstract may be used in abstracting and indexing databases, it
% should be self-contained (i.e., no numerical references) and substantive in
% nature, presenting concisely the objectives, methodology used, results
% obtained, and their significance. A list of up to six keywords should
% immediately follow. \end{abstract}

% See
% https://www.spiedigitallibrary.org/journals/journal-of-medical-imaging/volume-7/issue-01/010101/Integrating-structured-abstracts-in-the-Journal-of-Medical-Imaging/10.1117/1.JMI.7.1.010101.full
% for info on the structured abstract
%
% Purpose: This section presents the significance and aims, stating the broad
% impact and the rationale or motivation of the work, including potentially
% some background.
%
% Approach: This section briefly describes the materials and methods, including
% the statistical analyses, used in the research.
%
% Results: This section provides a core summary of the findings from the work,
% including study numbers, quantitative analyses, or discoveries.
%
% Conclusions: This section summarizes and interprets the approach and findings
% from the work, stating also the importance or impact of the findings.

% Purpose: Data-intensive modeling could provide insight on the broad
% variability in outcomes in spine surgery. Previous studies were limited to
% analysis of demographic and clinical characteristics. We report an analytic
% framework called “SpineCloud” that incorporates quantitative features
% extracted from perioperative images to predict spine surgery outcome.

% Approach: A retrospective study was conducted in which patient demographics,
% imaging, and outcome data were collected. Image features were automatically
% computed from perioperative CT. Postoperative 3- and 12-month functional and
% pain outcomes were analyzed in terms of improvement relative to the
% preoperative state. A boosted decision tree classifier was trained to predict
% outcome using demographic and image features as predictor variables.
% Predictions were computed based on SpineCloud and conventional demographic
% models, and features associated with poor outcome were identified from
% weighting terms evident in the boosted tree.

% Results Neither approach was predictive of 3- or 12-month outcomes based on
% preoperative data alone in the current, preliminary study. However,
% SpineCloud predictions incorporating image features obtained during and
% immediately following surgery (i.e., intraoperative and immediate
% postoperative images) exhibited significant improvement in area under the
% receiver operating characteristic (AUC): AUC=0.72 (CI95=0.59 to 0.83) at 3
% months and AUC=0.69 (CI95=0.55 to 0.82) at 12 months.

% Conclusions Predictive modeling of lumbar spine surgery outcomes was improved
% by incorporation of image-based features compared to analysis based on
% conventional demographic data. The SpineCloud framework could improve
% understanding of factors underlying outcome variability and warrants further
% investigation and validation in a larger patient cohort.
\begin{abstract}

@@

\section*{Approach}
We develop open-source software to efficiently compute RMT features from fMRI
images, and examine the cross-validated predictive potential of eigenvalue and
RMT-based features (``eigenfeatures'') with classic machine-learning
classifiers. We systematically vary pre-processing extent, normalization
procedures, RMT unfolding procedures, and feature selection, and compare the
impact of these analytic choices on the distributions of cross-validated
prediction performance for each combination of dataset binary classification
task, classifier, and feature. To deal with class imbalance, we use the
area under the receiver operating characteristic (AUROC) as the main
performance metric.

@@

\section*{Conclusions}
Eigenfeatures clearly have potential for understanding fMRI functional
connectivity in a wide variety of scenarios. However, the utility of these
features is strongly dependent on analytic decisions, suggesting caution when
interpreting past and future studies applying RMT to fMRI.

% This document shows the required format and appearance of a manuscript
% prepared for SPIE journals. It is prepared using LaTeX2e with the class file
% \texttt{spieman.cls}. Please note that the following journals require the use
% of structured abstracts in manuscript submissions: \textit{Neurophotonics},
% the \textit{Journal of Biomedical Optics}, and the \textit{Journal of Medical
% Imaging}. Structured abstracts are encouraged for the \textit{Journal of
% Micro/Nanolithography, MEMS, and MOEMS}. Guidelines are available on the
% journal website. Whether structured or single-paragraph, the abstract should
% be a summary of the paper and not an introduction. Because the abstract may
% be used in abstracting and indexing databases, it should be self-contained
% (i.e., no numerical references) and substantive in nature, presenting
% concisely the objectives, methodology used, results obtained, and their
% significance. A list of up to six keywords should immediately follow.
\end{abstract}

% Include a list of up to six keywords after the abstract
\keywords{random matrix, spectral rigidity, level number variance, fMRI, classification, machine-learning}

% Include email contact information for corresponding author
{\noindent \footnotesize\textbf{*}Jacob Levman,  \linkable{jlevman@stfx.ca} }

\begin{spacing}{2}   % use double spacing for rest of manuscript

\section{Introduction}
\label{sect:intro}  % \label{} allows reference to this section
% This document shows the format and appearance of a manuscript prepared for
% submission to an SPIE journal. Note that this template is only intended to be
% used as a guideline for author convenience. It is designed for optimum
% clarity and ease of reading for editors and reviewers, but the template does
% not reflect the final page layout of a published journal paper. Accepted
% papers are professionally typeset in XML according to the layout and design
% of the journal.

At its most rudimentary, RMT describes the expected behavior of the
eigenvalues—also often called levels—of a number of classes of random
matrices\cite{andersonIntroductionRandomMatrices2010,
akemannOxfordHandbookRandom2011, mehtaRandomMatrices2004}. A random matrix is,
simplifying somewhat, a matrix in which the entries are drawn from
independently and identically distributed (\textit{iid}) distributions. Though
first developed to describe the fluctuation of nuclei energy levels in quantum
physics \cite{mehtaRandomMatrices2004, guhrRandommatrixTheoriesQuantum1998a},
RMT has been shown have extremely broad potential.

In small-scale physical systems, RMT universalities have been observed in
quantum chaotic systems, complex nuclei, atoms, molecules and disordered
mesoscopic systems
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001},
and at larger scales, RMT has been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002} , social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids
\cite{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009}
and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}.
Within biological systems, RMT has also been used to successfully model aspects
of  amino acid functional relationships
\cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in epileptic seizures
\cite{osorioPhasesynchronizationRandommatrixBased2011}, and in protein-protein
interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}. RMT has also been used to guide
statistical decisions in principal components
analyses\cite{franklinParallelAnalysisMethod1995,
veraartDenoisingDiffusionMRI2016, ulfarssonDimensionEstimationNoisy2008}, and,
more recently, has provided insights into the behaviors of deep
neural networks\cite{martinImplicitSelfRegularizationDeep2021,
martinPredictingTrendsQuality2021}.

@@

For an imaging modality like fMRI, where changes in the
blood-oxygenation-level-dependent (BOLD) signals are related to neural
activity, RMT may be an ideal starting point, as each voxel time course (or
collection of such time sources - i.e. ROIs) can be considered an interacting
component of the system. Likewise, in functional connectivity
analyses, statistical relationships of the BOLD ROI time-courses are
investigated in the hope of gaining insights into brain functioning (see
\citenum{bastosTutorialReviewFunctional2016,
vandenheuvelExploringBrainNetwork2010} for reviews, but also
\citenum{nobleDecadeTestretestReliability2019} and
\citenum{lurieQuestionsControversiesStudy2020} for challenges facing functional
connectivity analyses). In this framework, each connection or correlation can
be considered a system component, and the eigenvalues of such a correlation
matrix can be examined from the perspective of RMT.

The earliest study taking this approach
demonstrated that spectra of the correlations between electroencephalographic
(EEG) signals closely resemble those of the Gaussian Orthogonal Ensemble (GOE)
\cite{sebaRandomMatrixAnalysis2003}. In fMRI, RMT has been used to
evaluate the quality of whole brain features extracted from fMRI data
\cite{voultsidouFeatureEvaluationFMRI2007,verganiRestingStateFMRI2019}, and in
diffusion MRI to aid in the selection of the number of
components to employ in principal-component reduction analysis and denoising
\cite{veraartDenoisingDiffusionMRI2016,verganiRestingStateFMRI2019,ulfarssonDimensionEstimationNoisy2008}.

@@

Typical reductions include the use of a ``seed'' voxel or collection of voxels
(see e.g. \citenum{joelRelationshipSeedbasedICAbased2011,
smithCharacterizingIndividualDifferences2014}): the mean signal over that ROI
is correlated with all other N ROIs of interest to generate a reduced
functional connectivity matrix (or image) with one correlation value at each
non-seed ROI. This sort of reduction reduces the functional connectivity to N
values, but necessarily biases the analysis to the seed ROI.

@@

Previous studies applying RMT to functional connectivity
data\cite{wangRandomMatrixTheory2016,
wangSpectralPropertiesTemporal2015a,matharooSpontaneousBackpainAlters2020,
guRandomMatrixTheory2020} took largely descriptive / explanatory approaches,
noting only differences in RMT across subgroups and/or conditions. For example
\citenum{wangRandomMatrixTheory2016} and \citenum{guRandomMatrixTheory2020} use
a Kolmogorov-Smirnov test, and \citenum{wangSpectralPropertiesTemporal2015a}
uses a t-test to provide evidence of differences in various RMT metrics.
However, statistically significant differences in a metric do not imply
practical significance or predictive utility
\cite{loWhySignificantVariables2015}, nor do they imply generalization or
replicability \cite{amrheinEarthFlat052017}. Cross-validation, by contrast,
attempts to more directly assess these properties
\cite{yarkoniChoosingPredictionExplanation2017}.

In addition, the previous papers do not provide publicly-accessible code to
reproduce results. The extraction of the spectral rigidity and level variance
is computationally-demanding and mathematically and algorithmically
non-trivial, and additionally requires an \textit{unfolding}
procedure\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}.
Unfolding is an exponential fitting procedure that involves a number of highly
subjective decisions regarding outliers and the flexibility of the fitting
function. These decisions are, unfortunately, often poorly or even entirely
undocumented, but are known to often dramatically impact RMT conclusions
\cite{abul-magdUnfoldingSpectrumChaotic2014,
abueleninSpectralUnfoldingChaotic2018, fossionRandommatrixSpectraTime2013,
abueleninEffectUnfoldingSpectral2012, moralesImprovedUnfoldingDetrending2011}.

On top of the flexibility in the implementation and application of RMT to
empirical data, there is also the flexibility introduced by the choices in the
complex preprocessing pipelines of fMRI\cite{parkerBenefitSliceTiming2019}.
These "researcher degrees of
freedom"\cite{simmonsFalsePositivePsychologyUndisclosed2011b} in combination
with the absence of reproducible code and data availability, and predominance
of descriptive and explanatory approaches, raise doubts about the basic robustness
and practical utility of RMT-based functional connectivity analyses.

@@

We selected fMRI data publicly available on the
\href{https://openneuro.org/}{OpenNeuro} platform (36). Selection criteria was
somewhat subjective, but we required (1), the dataset comprise 10+ human
subjects, (2) that all fMRI scans have the same spatial and temporal
resolutions, (3) that the fMRI images could be split into various classes for a
number of binary classification tasks, and (4), that it was reasonable to
classify each scan in its entirety, (i.e. neither task nor event timings
were needed to determine class membership, nor did any scan volumes need to be removed
to make the scan classifiable). This yielded seven datasets total.

\begin{table}[h!]
\caption{
    ID = Identifier for paper. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P horizontal / axial slices each with dimensions M x N.
}
\label{tab:data-dimensions}
\small
\centering
\begin{tabular}{ l c c c c c c }
\hline
\textbf{Name}    & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{n\_scans} \\
\hline
Aging     & 74 × 74 × 32   & 3.0 × 3.0 × 4.0 & 2.0  & 300 & 62  \\
Bilingual    & 100 × 100 × 72 & 1.8 × 1.8 × 1.8 & 0.88 & 823 & 90  \\
Depress   & 112 × 112 × 25 & 2.0 × 2.0 × 5.0 & 2.5  & 100 & 72  \\
Learn     & 64 × 64 × 36   & 3.0 × 3.0 × 3.0 & 2.0  & 195 & 432 \\
Osteo     & 64 × 64 × 36   & 3.4 × 3.4 × 3.0 & 2.5  & 300 & 74  \\
Park      & 80 × 80 × 43   & 3.0 × 3.0 × 3.0 & 2.4  & 149 & 552 \\
Attention & 128 × 128 × 70 & 1.5 × 1.5 × 1.5 & 3.0  & 300 & 90  \\
\hline
\end{tabular}
\end{table}

@@

\begin{table}[h!]
\caption{
    ID = Identifier for paper. Subgroup = name of subgroup. Subjects = Number of subjects.
    Task = fMRI task. ANT = Attention Network Task \cite{fanActivationAttentionalNetworks2005}
}
\label{tab:data-subjects}
\small
\centering
\begin{tabular}{ l r c c c }
\hline
\textbf{ID}  & \textbf{Subgroup}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
Aging   &  older             & 28      & resting-state        & 1 \\
        &  younger           & 34      & resting-state        & 1 \\
        &                    &         &                      &   \\
Bilingual  &  bilingual          & 59      & resting-state    & 1 \\
        &  monolingual        & 33      & resting-state       & 1 \\
        &                    &         &                      &   \\
Depress &  depression         & 51      & resting-state       & 1 \\
        &  control            & 21      & resting-state       & 1 \\
        &                    &         &                      &   \\
Learn   &  task              & 24      & learn image sketches & 16\\
        &  rest              & 24      & resting-state        & 2 \\
        &                    &         &                      &   \\
Osteo   &  duloxetine        & 19      & resting-state        & 1 \\
        &  pain              & 37      & resting-state        & 1 \\
        &  nopain            & 20      & resting-state        & 1 \\
        &                    &         &                      &   \\
Park    &  Parkinson's       & 25      & ANT                  & 12\\
        &  control           & 21      & ANT                  & 12\\
        &                    &         &                      &   \\
Attention&  vigilant           & 11     & resting-state       & 2 \\
        &  nonvigilant         & 11     & resting-state        & 2 \\
        &  trait-attentive     & 11    & resting-state        & 2 \\
        &  trait-non-attentive & 11    & resting-state        & 2 \\
        &  task-attentive     & 11     & resting-state        & 2 \\
        &  task-nonattentive  & 11     & resting-state        & 2 \\
\hline
\end{tabular}
\end{table}

@@

% Resting state fMRI scans were performed using a T2*-weighted EPI sequence with the following
% parameters: TR=875ms; TE=43.6ms; slice thickness=1.8mm; acquisition matrix=100 × 100; flip
% angle=55°; number of slices=72; field of view=180 × 180mm; voxel size=1.8 × 1.8 × 1.8mm;
% multi-band factor=8; volumes per run=823 (total scan time=12min). EPI scans were oriented
% parallel to the long axis of the hippocampus.

@@

% \subsubsection{Learning Dataset}

@@

% \subsubsection{Osteoarthritis Dataset}
The \textit{Osteo} dataset\cite{tetreaultBrainConnectivityPredicts2016}
includes whole-brain rs-fMRI scans of healthy, pain-free controls
(\code{nopain} condition), and individuals with knee osteoarthritis.
Osteoarthritic patients were treated for two weeks with either placebo
(\code{pain} condition) or duloxetine (\code{duloxetine} condition).
We use the three binary classification tasks \code{nopain v duloxetine},
\code{nopain v pain}, and \code{pain v duloxetine} for our analysis.

% \subsubsection{Parkinson's Dataset}
The \textit{Park} dataset\cite{madhyasthaDynamicConnectivityRest2015} includes
subjects with non-demented Parkinson's Disease (PD) and
healthy controls performed a number of repetitions of the Attention Network
Task (ANT\cite{fanActivationAttentionalNetworks2005}) during scans.
The classification task for this dataset is Parkinson's versus controls, i.e.
\code{park v ctrl}.

% \subsubsection{Attention Dataset}

@@

The \code{viglant v nonvigilant} task was formed based on self-report
questionnaire items involving
"vigilance"\cite{gorgolewskiCorrespondenceIndividualDifferences2014}. The
\code{trait\_attend v trait\_nonattend} task was formed using PANAS-X
\cite{watsonPANASXManualPositive1994} items related to self-reported
wakefulness and attention over the past weeks. Finally, the
\code{task\_attend v task\_nonattend} classification task was constructed
using the scores of the Conjunctive Continuous Performance Task
(CCPT\cite{shalevConjunctiveContinuousPerformance2011}). This is a behavioral
performance task which requires the subject to quickly and selectively respond
to only certain visual stimuli. Details allowing exact reproduction of these
splits are available in this paper's
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/preprocess/unify\_attention\_data.py}{source
code}.

@@

We limited preprocessing to a simple pipeline of, in order: (1) brain
extraction, (2) slice timing correction, (3), motion correction, and (4),
registration to the MNI152
\cite{fonovUnbiasedAverageAgeappropriate2011,fonovUnbiasedNonlinearAverage2009}
2mm isotropic, asymmetric template version C made available through
\href{https://www.templateflow.org/}{TemplateFlow}
\cite{ciricTemplateFlowFAIRsharingMultiscale2021}.

>>>>

\title{Random Matrix Theory Tools for the Predictive Analysis of Functional Magnetic Resonance Imaging Examinations}

\author[a]{Derek Berger}
\author[b,c]{Gurpreet S. Matharoo}
\author[a*,d,e,f]{Jacob Levman}

\affil[a]{St. Francis Xavier University, Department of Computer Science, 4130 University Avenue, Antigonish, Nova Scotia, Canada, B2G 2W5}
\affil[b]{St. Francis Xavier University, ACENET, 4130 University Avenue, Antigonish, Nova Scotia, Canada, B2G 2W5}

\affil[c]{St. Francis Xavier University, Department of Physics, 4130 University Avenue, Antigonish, Nova Scotia, Canada, B2G 2W5}

\affil[d]{Martinos Center for Biomedical Imaging, 149 Thirteenth Street, Suite 2301, Charlestown, Massachusetts, United States, 02129}
\affil[e]{Harvard Medical School, Department of Radiology, 25 Shattuck Street Boston, Massachusetts 02115, United States, 02129}
\affil[f]{Nova Scotia Health Authority, Research Affiliate, Nova Scotia, Canada}

@@

\begin{abstract}

@@

\section*{Approach}
We develop open-source software to efficiently compute RMT features from fMRI
images, and examine the cross-validated predictive potential of eigenvalue and
RMT-based features (``eigenfeatures'') with classic machine-learning
classifiers. We systematically vary pre-processing extent, normalization
procedures, RMT unfolding procedures, and feature selection, and compare the
impact of these analytic choices on the distributions of cross-validated
prediction performance for each combination of dataset binary classification
task, classifier, and feature. To deal with class imbalance, we use the
area under the receiver operating characteristic curve (AUROC) as the main
performance metric.

@@

\section*{Conclusions}
Eigenfeatures clearly have potential for understanding fMRI functional
connectivity in a wide variety of scenarios. The utility of these features is
strongly dependent on analytic decisions, suggesting caution when interpreting
past and future studies applying RMT to fMRI. However, this study demonstrates
that the inclusion of RMT statistics in fMRI investigations could improve
prediction performances across a wide variety of phenomena.

\end{abstract}

\keywords{random matrix, spectral rigidity, level number variance, fMRI, classification, machine-learning}

{
\noindent
\footnotesize
\textbf{*}Jacob Levman,  \linkable{jlevman@stfx.ca}
(primary), \linkable{jlevman@mgh.harvard.edu}
}

\begin{spacing}{2}

\section{Introduction}
\label{sect:intro}

At its most rudimentary, RMT describes the expected behavior of the
eigenvalues—also often called levels—of a number of classes of random
matrices\cite{andersonIntroductionRandomMatrices2010,
akemannOxfordHandbookRandom2011, mehtaRandomMatrices2004}. A random matrix is,
simplifying somewhat, a matrix in which the entries are drawn from
independently and identically distributed (\textit{iid}) distributions. Though
first developed to describe the fluctuation of nuclei energy levels in quantum
physics \cite{mehtaRandomMatrices2004, guhrRandommatrixTheoriesQuantum1998a},
RMT has been shown to have extremely broad potential. In small-scale physical
systems, RMT universalities have been observed in quantum chaotic systems,
complex nuclei, atoms, molecules and disordered mesoscopic systems
\cite{guhrRandommatrixTheoriesQuantum1998a, mehtaRandomMatrices2004,
brodyRandommatrixPhysicsSpectrum1981, beenakkerRandommatrixTheoryQuantum1997,
bohigasHigherOrderCorrelationsSpectra1985, wintgenLevelStatisticsQuantized1988,
pandeySkewOrthogonalPolynomialsUniversality2001}, and at larger scales, RMT has
been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002}, social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids\cite{sastrySpectralStatisticsInstantaneous2001,
matharooSpectralStatisticsQuenched2009} and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,
matharooVibrationalSpectraAmorphous2005,
matharooUniversalityVibrationalSpectra2008}. Within biological systems, RMT has
also been used to successfully model aspects of  amino acid functional
relationships \cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in
epileptic seizures \cite{osorioPhasesynchronizationRandommatrixBased2011}, and
in protein-protein interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}. RMT has also been used to guide
statistical decisions in principal components
analyses\cite{franklinParallelAnalysisMethod1995,
veraartDenoisingDiffusionMRI2016, ulfarssonDimensionEstimationNoisy2008}, and,
more recently, has provided insights into the behaviors of deep neural
networks\cite{martinImplicitSelfRegularizationDeep2021,
martinPredictingTrendsQuality2021}.

@@

For an imaging modality like fMRI, where changes in the
blood-oxygenation-level-dependent (BOLD) signals are related to neural
activity, RMT may be an ideal starting point, as each voxel time course (or
collection of such time sources, i.e. ROIs) can be considered an interacting
component of the system. Likewise, in functional connectivity
analyses, statistical relationships of the BOLD ROI time-courses are
investigated in the hope of gaining insights into brain function (see
\citenum{bastosTutorialReviewFunctional2016,
vandenheuvelExploringBrainNetwork2010} for reviews, but also
\citenum{nobleDecadeTestretestReliability2019} and
\citenum{lurieQuestionsControversiesStudy2020} for challenges facing functional
connectivity analyses). In this framework, each connection or correlation can
be considered a system component, and the eigenvalues of such a correlation
matrix can be examined from the perspective of RMT.

The earliest study taking this approach demonstrated that spectra of the
correlations between electroencephalographic (EEG) signals closely resemble
those of the Gaussian Orthogonal Ensemble (GOE)
\cite{sebaRandomMatrixAnalysis2003}. In fMRI, RMT has been used to evaluate the
quality of whole brain features extracted from fMRI data
\cite{voultsidouFeatureEvaluationFMRI2007, verganiRestingStateFMRI2019}, and in
diffusion MRI to aid in the selection of the number of components to employ in
principal-component reduction analysis and denoising
\cite{veraartDenoisingDiffusionMRI2016, verganiRestingStateFMRI2019,
ulfarssonDimensionEstimationNoisy2008}.

@@

Typical reductions include the use of a ``seed'' voxel or collection of voxels
(see e.g. \citenum{joelRelationshipSeedbasedICAbased2011,
smithCharacterizingIndividualDifferences2014}): the mean signal over that ROI
is correlated with all other \(N\) ROIs of interest to generate a reduced
functional connectivity matrix (or image) with one correlation value at each
non-seed ROI. This sort of reduction reduces the functional connectivity to
\(N\) values, but necessarily biases the analysis to the seed ROI.

@@

Previous studies applying RMT to functional connectivity
data\cite{wangRandomMatrixTheory2016, wangSpectralPropertiesTemporal2015a,
matharooSpontaneousBackpainAlters2020, guRandomMatrixTheory2020} took largely
descriptive / explanatory approaches, noting only differences in RMT across
subgroups and/or conditions. For example Wang et al.
(2016)\cite{wangRandomMatrixTheory2016} and Gu et al.
(2020)\cite{guRandomMatrixTheory2020} use a Kolmogorov-Smirnov test, and Wang
et al. (2015)\cite{wangSpectralPropertiesTemporal2015a} uses a t-test to
provide evidence of differences in various RMT metrics. However, statistically
significant differences in a metric do not necessarily imply practical
significance or predictive utility \cite{loWhySignificantVariables2015}, nor do
they imply generalization or replicability \cite{amrheinEarthFlat052017}.
Cross-validation, by contrast, attempts to more directly assess these
properties \cite{yarkoniChoosingPredictionExplanation2017}.

In addition, the previous papers do not provide publicly-accessible code to
reproduce results. The extraction of the spectral rigidity and level variance
is computationally-demanding and mathematically and algorithmically
non-trivial, and additionally requires an \textit{unfolding}
procedure\cite{guhrRandommatrixTheoriesQuantum1998a, mehtaRandomMatrices2004}.
Unfolding is an exponential fitting procedure that involves a number of highly
subjective decisions regarding outliers and the flexibility of the fitting
function. These decisions are, unfortunately, often poorly or even entirely
undocumented, but are known to often dramatically impact RMT conclusions
\cite{abul-magdUnfoldingSpectrumChaotic2014,
abueleninSpectralUnfoldingChaotic2018, fossionRandommatrixSpectraTime2013,
abueleninEffectUnfoldingSpectral2012, moralesImprovedUnfoldingDetrending2011}.

In addition to the flexibility in the implementation and application of RMT to
empirical data, there is also flexibility introduced by analytic choices made
in the complex preprocessing pipelines of
fMRI\cite{parkerBenefitSliceTiming2019}. These "researcher degrees of
freedom"\cite{simmonsFalsePositivePsychologyUndisclosed2011b}, in combination
with the absence of reproducible code and data availability, and predominance
of descriptive and explanatory approaches, raise doubts about the basic
robustness and practical utility of RMT-based functional connectivity analyses.

@@

We selected fMRI data publicly available on the
\href{https://openneuro.org/}{OpenNeuro}
platform\cite{markiewiczOpenNeuroResourceSharing2021}. Selection criteria was
somewhat subjective, but we required each dataset to (1) comprise 10 or more
human subjects, (2) have all fMRI scans have the same spatial and temporal
resolutions, (3) have fMRI images that can be split into various classes for a
number of binary classification tasks, and (4), allow reasonably classifying
each scan in its entirety (i.e. neither task nor event timings are needed to
determine class membership, nor do any scan volumes need to be removed to make
the scan classifiable). This yielded seven datasets total.

\begin{table}[h!]
\caption{
    \label{tab:data-dimensions}
    Included fMRI dataset details.
    Name = Identifier for paper.
    Dimensions listed as M x N x P, indicate P axial slices each with dimensions M x N.
    TR = Time of Repetition (seconds).
    n\_scans = Total number of images in dataset.
}
\small
\centering
\begin{tabular}{ l c c c c c c }
\hline
\textbf{Name}    & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{n\_scans} \\
\hline
Aging     & 74 × 74 × 32   & 3.0 × 3.0 × 4.0 & 2.0  & 300 & 62  \\
Bilingual    & 100 × 100 × 72 & 1.8 × 1.8 × 1.8 & 0.88 & 823 & 90  \\
Depress   & 112 × 112 × 25 & 2.0 × 2.0 × 5.0 & 2.5  & 100 & 72  \\
Learn     & 64 × 64 × 36   & 3.0 × 3.0 × 3.0 & 2.0  & 195 & 432 \\
Osteo     & 64 × 64 × 36   & 3.4 × 3.4 × 3.0 & 2.5  & 300 & 74  \\
Park      & 80 × 80 × 43   & 3.0 × 3.0 × 3.0 & 2.4  & 149 & 552 \\
Attention & 128 × 128 × 70 & 1.5 × 1.5 × 1.5 & 3.0  & 300 & 90  \\
\hline
\end{tabular}
\end{table}

@@

\begin{table}[h!]
\caption{
    \label{tab:data-subjects}
    Sizes and other details for classification task subgroups.
    ID = Identifier for paper. Subgroup = name of subgroup used in classification task. Subjects = Number of subjects.
    Task = fMRI task. ANT = Attention Network Task \cite{fanActivationAttentionalNetworks2005}
}
\small
\centering
\begin{tabular}{ l r c c c }
\hline
\textbf{ID}  & \textbf{Subgroup}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
Aging   &  older             & 28      & resting-state        & 1 \\
        &  younger           & 34      & resting-state        & 1 \\
        &                    &         &                      &   \\
Bilingual  &  bilingual          & 59      & resting-state    & 1 \\
        &  monolingual        & 33      & resting-state       & 1 \\
        &                    &         &                      &   \\
Depress &  depression         & 51      & resting-state       & 1 \\
        &  control            & 21      & resting-state       & 1 \\
        &                    &         &                      &   \\
Learn   &  task              & 24      & learn image sketches & 16\\
        &  rest              & 24      & resting-state        & 2 \\
        &                    &         &                      &   \\
Osteo   &  duloxetine        & 19      & resting-state        & 1 \\
        &  pain              & 37      & resting-state        & 1 \\
        &  nopain            & 20      & resting-state        & 1 \\
        &                    &         &                      &   \\
Park    &  Parkinson's       & 25      & ANT                  & 12\\
        &  control           & 21      & ANT                  & 12\\
        &                    &         &                      &   \\
Attention&  vigilant           & 11     & resting-state       & 2 \\
        &  nonvigilant         & 11     & resting-state        & 2 \\
        &  trait-attentive     & 11    & resting-state        & 2 \\
        &  trait-non-attentive & 11    & resting-state        & 2 \\
        &  task-attentive     & 11     & resting-state        & 2 \\
        &  task-nonattentive  & 11     & resting-state        & 2 \\
\hline
\end{tabular}
\end{table}

@@

% Resting state fMRI scans were performed using a T2*-weighted EPI sequence
% with the following parameters: TR=875ms; TE=43.6ms; slice thickness=1.8mm;
% acquisition matrix=100 × 100; flip angle=55°; number of slices=72; field of
% view=180 × 180mm; voxel size=1.8 × 1.8 × 1.8mm; multi-band factor=8; volumes
% per run=823 (total scan time=12min). EPI scans were oriented parallel to the
% long axis of the hippocampus.

@@

@@

The \textit{Osteo} dataset\cite{tetreaultBrainConnectivityPredicts2016}
includes whole-brain rs-fMRI scans of healthy, pain-free controls
(\code{nopain} condition), and individuals with knee osteoarthritis.
Osteoarthritic patients were treated for two weeks with either placebo
(\code{pain} condition) or duloxetine (\code{duloxetine} condition).
We use the three binary classification tasks \code{nopain v duloxetine},
\code{nopain v pain}, and \code{pain v duloxetine} for our analysis.

The \textit{Park} dataset\cite{madhyasthaDynamicConnectivityRest2015} includes
subjects with non-demented Parkinson's Disease (PD) and
healthy controls performed a number of repetitions of the Attention Network
Task (ANT\cite{fanActivationAttentionalNetworks2005}) during scans.
The classification task for this dataset is Parkinson's versus controls, i.e.
\code{park v ctrl}.

@@

The \code{vigilant v nonvigilant} task was formed based on self-report
questionnaire items involving
"vigilance"\cite{gorgolewskiCorrespondenceIndividualDifferences2014}. The
\code{trait\_attend v trait\_nonattend} task was formed using PANAS-X
\cite{watsonPANASXManualPositive1994} items related to self-reported
wakefulness and attention over the past weeks. Finally, the
\code{task\_attend v task\_nonattend} classification task was constructed
using the scores of the Conjunctive Continuous Performance Task
(CCPT\cite{shalevConjunctiveContinuousPerformance2011}). This is a behavioral
performance task which requires the subject to quickly and selectively respond
to only certain visual stimuli. Details allowing exact reproduction of these
splits are available in this paper's
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/preprocess/unify\_attention\_data.py}{source
code}.

@@

We limited preprocessing to a simple pipeline of, in order: (1) brain
extraction, (2) slice timing correction, (3), motion correction, and (4),
registration to the MNI152
\cite{fonovUnbiasedAverageAgeappropriate2011, fonovUnbiasedNonlinearAverage2009}
2mm isotropic, asymmetric template version C made available through
\href{https://www.templateflow.org/}{TemplateFlow}
\cite{ciricTemplateFlowFAIRsharingMultiscale2021}.

Brain extraction was performed first with BET
\cite{smithFastRobustAutomated2002}, and then with
ANTs\cite{avantsReproducibleEvaluationANTs2011} to clean up residue left behind
by BET. BET results were visually inspected for each image in each dataset, and
BET fractional anisotropy values were modified to ensure brain extraction was
acceptable. Motion-correction was performed with FSL's MCFLIRT
\cite{jenkinsonImprovedOptimizationRobust2002}, and slice time correction
with FSL's slicetimer\cite{jenkinsonFSL2012} tool. Finally, functional
images were registered directly to the MNI152 template via
ANTs\cite{avantsReproducibleEvaluationANTs2011}.
